---
title: "An Overview of Functional Analysis"
output:
        html_document:
            includes:
                in_header: header.html
            theme: paper
            toc: true
            toc_depth: 4
            toc_float: true
            number_sections: false
            fig_caption: false
            markdown: GFM
            code_download: false
---

<style type="text/css">

body, td {
   font-size: 16px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}

</style>

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(lang = c("r", "markdown", "bash"), position = c("top", "right"))
```

```{r, 'chunk_options', include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```

```{r, include = FALSE}
source("../bin/set_values.R")
```

# Objectives

1. Consider why we do functional analysis.
2. Walk through an example overrepresentation analysis (ORA).
3. Introduce the three high-level approaches to functional analysis
4. Outline the topics for the workshop.


# Why Functional analysis 

![DEG analysis flowchart]()

Consider this plot:
![DEG volcano plot]()

How would you interpret it?

1) Fold change: Magnitude of the effect
2) p-value: Statistical significance
2a) FDR/Adjusted p-value: significance adjusted for multiple test correction

What does it show in terms of the biology of the model system? In terms of the
research question?

Research question may involve specific genes, but is invariably rooted in 
biological systems. Functional analysis can bridge this gap.

Imagine we constructed ths plot based on the volcano plot above:

!(Curiously simple enrichment plot of five reference gene sets)[]

Which pathways are enriched? Are they statistically significant?



Let's consider a single reference gene set in a very small universe


Visualizing enrichment
!(Venn diagrams)[]
((query set) background set) = background proportion
((query set) reference set) = reference proportion


Diagram with gene sets
query set
reference set


2x2 table
significant in category proportions = a/(a+b)
gs in background proportions = (a+c) / N
fold enrichment =  significant in gs / gs in background = observed / expected
= 2.667
The pathway is ~2.67× overrepresented among your significant genes relative to the background.

>1 enrichment, <1 depletion
fisher's exact test / hypergeometric 
p = 0.014

FOLD: odds ratio
OR = odds a (gene is in gs | significant) / odds(gene in gs| not significant)
OR = 5; sig genes 5 times more likely to be in gs than background genes
>1 positive assocation (<1 negative association)


Let's define gene set.


Now expand to many gene sets.

(And now with FDR correction)


# ORA: Pros & cons:

## ORA Pros

- Analysis is relatively simple to execute.
- Interpretation is straightforward.
- Supports a wide range of biological processes and model systems.
- Significant results are a robust indicator of biological activity.
  
## ORA Cons

  - Requires a background set of genes. Generally straightforward for RNA-Seq
    (transcriptome) or microarray, more nuanced for proteins.
  - Requires a subset of diffex genes, usually partitioned with arbitrary cutoff value. 
  - Treats all genes the same; doesn't care whether the gene is over or under 
    or by how much. (That's why the inputs are not ranked or ordered - they are 
    simply sets of genes.)
  - Too few diffex genes or too many diffex genes will miss true associations.


# Common functional analysis approaches

- ORA 
- FCS / GSEA
- TP

Similar steps / Similar inputs / Similar outputs
Our focus / path in this workshop
An example builk RNA-Seq experiment.

  
Note that we are focusing on gene expression, but functional analysis can be applied to
many experiment designs and analytes including non-coding RNA, protein 
expression, metabolites, and DNA methylation.
   
# Summary 

1. Functional analysis is a useful extension of a differential expression (DE) analysis:

  - Adding biological context can connect observed DE patterns with the research 
    question. 
  
  - Adding biological context can reduce complexity: Functional analysis starts 
    with hundreds to thousands of genes and emits dozens of biological processes 
    or functional modules. This can greatly simplify interpretation. 
    
  - Aggregating the effects of individual genes into processes or functional 
    modules reveals coordinated changes. 
    
  - Functional analysis results are simpler to generalize and compare across 
    individual DE experiments.


2. There are many tools, but three common approaches to functional analysis:

  - Overreprestation Analysis (ORA)
  - Functional Class Scoring (FCS) / Gene Set Enrichment Analysis (GSEA)
  - Topology / Pathway analysis (TP)
  
3. The details of each of these approaches are distinct, but they share similar 
   high-level steps and *outputs*:

  - Start with a collection of genes, aka the *query genes*.
  - Compare the query genes with other gene sets
  - Determine *enrichment*, *statistical significance*, *adjusted significance*.


4. Understanding ORA and FCS provides a solid foundation for running and 
   interpreting functional analysis so the workshop will focus on these
   two approaches along with some discussion of key reference databases.
   We will start with an exploration of ORA on bulk RNA-Seq data from a specific 
   mouse experiment.

<hr/>

## References

1. https://bioinformatics.ccr.cancer.gov/btep/pathways-and-gene-sets-what-is-functional-enrichment-analysis/
2. https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1002375#pcbi-1002375-t001

| [Previous lesson](workshop-intro.html) | [Top of this lesson](#top) | [Next lesson](more-stuff.html) |
| :--- | :----: | ---: |