---
title: "Functional Analysis and GSEA"
output:
        html_document:
            includes:
                in_header: header.html
            theme: paper
            toc: true
            toc_depth: 4
            toc_float: true
            number_sections: false
            fig_caption: false
            markdown: GFM
            code_download: false
---

<style type="text/css">

body, td {
   font-size: 16px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}

</style>

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(lang = c("r", "markdown", "bash"), position = c("top", "right"))
```

```{r, 'chunk_options', include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```

```{r, include = FALSE}
source("../bin/set_values.R")
```

# Objectives

1. Review top level goals of functional analysis
2. Review remianing topics for the workshop.
3. Walk through FCS/GSEA's approach to functional analysis.

# Common functional analysis approaches

- ORA 
- FCS / GSEA
- TP

Similar steps / Similar inputs / Similar outputs
Our focus / path in this workshop

# ORA Review

Schematic

## ORA Pros

- Analysis is relatively simple to execute.
- Interpretation is straightforward.
- Supports a wide range of biological processes and model systems.
- Significant results are a robust indicator of biological activity.
  
## ORA Cons

  - Requires a background set of genes. Generally straightforward for RNA-Seq
    (transcriptome) or microarray, more nuanced for proteins.
  - Requires a subset of diffex genes, usually partitioned with arbitrary cutoff 
    value. Too few diffex genes or too many diffex genes will miss real 
    functional relationships.

  - Treats all genes the same; doesn't care whether the gene is over or under 
    or by how much. (That's why the inputs are not ranked or ordered - they are 
    simply sets of genes.)


# FCS/GSEA

An example sc RNA-Seq experiment.



# Summary 

1. 
2. 
3.
4.
<hr/>

## References

1. 
2.

| [Previous lesson](07A-advanced-visualizations.html) | [Top of this lesson](#top) | [Next lesson](06-WebGestaltRGSEA.html) |
| :--- | :----: | ---: |